Evaluating Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe using radiocarbon data

### 2019-03-31 article/02_Materials_and_Methods.Rmd
started to incorporate the new arrangements of Figure 6 and Partial Mantel test into the text

From these time series another set of time series with a measure of cultural distance for each region-region relationship can be derived (Figure \ref{sed_region_matrix}) -- though with a reasonable amount of caution due to the partially low number of data available to reconstruct the developments in several regions and time slots. Cultural distance is defined here as the squared Euclidean distance (SED) between the trait proportions of two different regions at a certain point in time [@neimanStylisticVariationEvolutionary1995]. SED is a sensible metric here, because it is relatively simple, works with only a few assumptions and emphasises differences through squaring [@nakoinzFingerprintingIronAge2014].

@@

To test for correlation between the distances based on the variables *burial type* and *burial construction*, the time frame of interest was divided into sections of 200 years. For each of the $1400/200 = 7$ section and each of the two variables, a matrix of the mean SED over time was calculated. The two matrices of each section were tested for correlation with the Mantel test, which uses permutations of potentially correlating matrices to determine how strongly random reordering can alter an underlying correlation coefficient [@mantelDetectionDiseaseClustering1967]. Since both distances compared here, *burial type* and *burial construction* distance, are cardinal scale variables, the Pearson correlation coefficient [@pearsonNotesRegressionInheritance1895] can be applied as a base measure for the Mantel test (Table \ref{tab:dacta} C).

Defining a measure of spatial distance for the above established main units of observation, the macro-regions, is only possible in a highly simplified way. A continuous distance metric such as least-cost path distance cannot be employed on this scale, but it is possible to distinguish ordinal distance classes. For the eight regions, four different categories of spatial distance were defined (Figure \ref{map_regions} B). The test for correlation between spatial and cultural distance works similar to the one between the two cultural distances above, except that the Mantel test has to be applied with Spearman's rank correlation coefficient [@spearmanProofMeasurementAssociation1904] here, because of the different data levels of the measurements (Table \ref{tab:dacta} AB).

@@

```{r dactb, echo=FALSE, tidy=FALSE}
load("distance_and_correlation_table_B.RData")
distance_and_correlation_table_B %>%
  dplyr::mutate_at(
    dplyr::vars(dplyr::matches("Correlation")),
    list(~kableExtra::cell_spec(., "latex", color = ifelse(. < 0, "red", "black")))
  ) %>%
  dplyr::mutate_at(
    dplyr::vars(dplyr::matches("p-Value")),
    list(~kableExtra::cell_spec(., "latex", bold = . < 0.1))
  ) %>%
knitr::kable(
  "latex",
  caption = "Like Table \\ref{tab:dacta}. \\textbf{C} shows the Mantel test results for the two variables if compared directly, \\textbf{D} the results of the Partial Mantel test considering spatial distance as a second independent variable.",
  booktabs = T,
  linesep = "\\addlinespace \\hline \\addlinespace",
  escape = F
) %>%
  kableExtra::add_header_above(
    c(" ", "variable correlation" = 2, "variable correlation with spatial correction" = 2)
  ) %>%
  kableExtra::add_header_above(
    c(" ", "C" = 2, "D" = 2),
    bold = T, line = F
  ) %>%
  kableExtra::kable_styling(
    latex_options = c("hold_position"),
    font_size = 9,
    full_width = T
  ) %>%
  kableExtra::column_spec(c(1), width = "4cm") %>%
  gsub("table", "table*", .)
```

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{correlation_time_series.jpeg}
\caption{\label{simulation}Development of correlation between cultural and spatial distance. The results for the burial customs data (Table \ref{tab:dacta} AB) are displayed with a solid (\textit{burial type}) and dashed (\textit{burial construction}) line. The horizontal line marks the correlation coefficient of zero, where neither positive nor negative correlation can be assumed.}
\end{figure}

>>>>

From these time series another set of time series with a measure of cultural distance for each region-region relationship can be derived (Figure \ref{sed_region_matrix}) -- though with a reasonable amount of caution due to the partially low number of data available to reconstruct the developments in several regions and time slots. Cultural distance is defined here as the squared Euclidean distance (SED) between the trait proportions of two different regions at a certain point in time [@neimanStylisticVariationEvolutionary1995]. SED is a sensible metric in this context, because it is relatively simple, works with only a few assumptions and emphasises differences through squaring [@nakoinzFingerprintingIronAge2014].

@@

To test for correlation between the distances based on the variables *burial type* and *burial construction*, the time frame of interest was divided into sections of 200 years. For each of the $1400/200 = 7$ section and each of the two variables, a matrix of the mean SED over time was calculated. The two matrices of each section were tested for correlation with the *Mantel test*, which uses permutations of potentially correlating matrices to determine how strongly random reordering can alter an underlying correlation coefficient [@mantelDetectionDiseaseClustering1967]. Since both distances compared here, *burial type* and *burial construction* distance, are cardinal scale variables, the Pearson correlation coefficient [@pearsonNotesRegressionInheritance1895] can be applied as a base measure for the Mantel test (Table \ref{tab:dactb} C).

Defining a measure of spatial distance for the above established main units of observation, the macro-regions, is only possible in a highly simplified way. A continuous distance metric such as least-cost path distance cannot be employed on this scale, but it is possible to distinguish ordinal distance classes. For the eight regions, four different categories of spatial distance were defined (Figure \ref{map_regions} B). The test for correlation between spatial and cultural distance works similar to the one between the two cultural distances above, except that the Mantel test has to be applied with Spearman's rank correlation coefficient [@spearmanProofMeasurementAssociation1904], because of the different data levels of the measurements (Table \ref{tab:dacta} AB).

With a *Partial Mantel test* the spatial distance can also be considered for the direct comparison of the SED networks of *burial type* and *burial construction* (Table \ref{tab:dactb} D). While the Mantel test only allows a comparison among two variables, the Partial Mantel test allows one to be made among two variables while controlling for the third. In this case again Spearman's rank correlation coefficient has to be applied.

Figure \ref{correlationtimeseries} shows the development of the correlation coefficient values obtained via Mantel statistics (Table \ref{tab:dacta} & \ref{tab:dactb}) during the succession of 200 years time slots.

@@

```{r dactb, echo=FALSE, tidy=FALSE}
load("distance_and_correlation_table_B.RData")
distance_and_correlation_table_B %>%
  dplyr::mutate_at(
    dplyr::vars(dplyr::matches("Correlation")),
    list(~kableExtra::cell_spec(., "latex", color = ifelse(. < 0, "red", "black")))
  ) %>%
  dplyr::mutate_at(
    dplyr::vars(dplyr::matches("p-Value")),
    list(~kableExtra::cell_spec(., "latex", bold = . < 0.1))
  ) %>%
knitr::kable(
  "latex",
  caption = "Like Table \\ref{tab:dacta}. \\textbf{C} shows the Mantel test results for the two variables if compared directly, \\textbf{D} the results of the Partial Mantel test considering and correcting for spatial distance as a second independent variable.",
  booktabs = T,
  linesep = "\\addlinespace \\hline \\addlinespace",
  escape = F
) %>%
  kableExtra::add_header_above(
    c(" ", "variable correlation" = 2, "variable correlation with spatial correction" = 2)
  ) %>%
  kableExtra::add_header_above(
    c(" ", "C" = 2, "D" = 2),
    bold = T, line = F
  ) %>%
  kableExtra::kable_styling(
    latex_options = c("hold_position"),
    font_size = 9,
    full_width = T
  ) %>%
  kableExtra::column_spec(c(1), width = "4cm") %>%
  gsub("table", "table*", .)
```

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{correlation_time_series.jpeg}
\caption{\label{correlationtimeseries}Development of correlation between cultural and spatial distance. The results for the burial customs data (Table \ref{tab:dacta} AB) are displayed with a solid (\textit{burial type}) and dashed (\textit{burial construction}) line. The horizontal line marks the correlation coefficient of zero, where neither positive nor negative correlation can be assumed.}
\end{figure}

### 2019-03-31 article/04_Discussion.Rmd
started to incorporate the new arrangements of Figure 6 and Partial Mantel test into the text

Against this background, the observation that there is no permanent, significant correlation either between the variables or with the spatial distance is all the more astonishing (Table \ref{tab:dacta}). Only for small periods of time this correlation is visible in the Radon-B data. The *burial construction* distance shows a higher degree of correlation with spatial distance than the *burial type* distance, which even seems to behave in an opposing manner: negative correlation. If one assumes the reconstructed timelines of cultural similarity intensity to be correct, than one has to explain a rise (between 2200 and 2000 calBC) and a decline (after 2000 calBC) of cultural-spatial-distance correlation. There are several reasons why any interpretation of this observation is difficult so far:

>>>>

Against this background, the observation that there is no permanent, significant correlation either between the variables or with the spatial distance is all the more astonishing (Table \ref{tab:dacta}&\ref{tab:dacta}). Only for small periods of time this correlation is visible in the Radon-B data. The *burial construction* distance shows a higher degree of correlation with spatial distance than the *burial type* distance, which even seems to behave in an opposing manner: negative correlation. If one assumes the reconstructed timelines of cultural similarity intensity to be correct, than one has to explain a rise (between 2200 and 2000 calBC) and a decline (after 2000 calBC) of cultural-spatial-distance correlation. There are several reasons why any interpretation of this observation is difficult so far:
